Tails or Types? A Critical Systematic Review of Taxometric Studies in Neurodevelopmental Disorders

### 2025-07-20 Paper/Paper.qmd
paper update

Taxometric analysis is not without limitations, however. Commonly cited threats to validity of taxometric conclusions include poor indicator validity, nuisance covariance, low base rate, and artificial admixture \[...\]. Poor indicator validity occurs when taxometric analysis is run on variables that present little mean differences between individuals with and without the target disorder (i.e., putatively, the taxon and the complement). Generally, it is recommended that Cohen's d ≫ 1 on each indicator. Nuisance covariance means that there is strong residual correlations after accounting for taxa (i.e., both within the taxon and the complement), thus reflecting the use of redundant indicators; residual r \> 0.40 is often regarded as problematic. Low base rate means that individuals with the taxonic condition are too few in the sample for taxometric analysis to correctly identify them; n \< 30 is often regarded as problematic. This is obviously a problem when the diagnostic category is rare (e.g., 1-2% of the population). All of these threats lead to the risk of failing to detect taxa where they exist. On the other side, artificial admixture consists of collecting participants using alternate recruitment channels. This is often (incorrectly) used to compensate for the low base rate, as many individuals with a diagnosis are recruited separately from the rest of the sample to increase the number of observations with the putative taxon. The resulting sample is an admixture of individuals all of whom might originally come from the same dimensional population, yet it gives the appearance that multiple subgroups are there.

In addition, many studies on autism have drawn categorical conclusions from data that likely suffer from artificial admixture—mixing separately recruited clinical and control samples—which is known to spuriously generate taxonic results even when the true latent structure is dimensional @grove1991; @ruscio2004clarifying; @toffalini2024airipa. While a handful of studies (e.g., @frazier2023categorical) suggest that autism may exhibit a categorical structure, these conclusions warrant skepticism given that they are typically based on admixed samples and rarely report key methodological parameters such as indicator validity or within-group covariance.

## Aims of the Present Review

# Methods

## Search Strategy and Study Selection

## Coding

very strong indicator skewness (for example due to using scales that only measure very rare and seriously problematic behaviors) might generate evidence for taxa where none exist.

# Results

@@

The coexistence of nuisance covariance and/or poor indicator validity is practically unavoidable when true latent structure is dimensional. This is because, for an indicator to largely separate individuals with vs without a diagnosis, it must also be strongly related to the latent dimension, thus leading to strong nuisance covariance with any other indicator. But in fact, such empasse might directly reveal that the true structure of the putative disorder is dimensional.

>>>>

Taxometric analysis is not without limitations, however, some of which are shared with cluster analysis. Commonly cited threats to validity of taxometric conclusions include poor indicator validity, nuisance covariance, low base rate, and artificial admixture \[...\]. Poor indicator validity occurs when taxometric analysis is run on variables that present little mean differences between individuals with and without the target disorder (i.e., putatively, the taxon and the complement). Generally, it is recommended that Cohen's d ≫ 1 on each indicator. Nuisance covariance means that there is strong residual correlations after accounting for taxa (i.e., both within the taxon and the complement), thus reflecting the use of redundant indicators; residual r \> 0.40 is often regarded as problematic. Low base rate means that individuals with the taxonic condition are too few in the sample for taxometric analysis to correctly identify them; n \< 30 is often regarded as problematic. This is obviously a problem when the diagnostic category is rare (e.g., 1-2% of the population). Finally, artificial admixture consists of mixing participants selected or recruited using alternate criteria. This is often (incorrectly) used to compensate for the low base rate, with many individuals with a diagnosis being recruited separately from the rest of the sample to increase the number of observations with the putative taxon. Poor indicator validity and low base rate risk masking detection of taxa, while strong nuisance covariance and artificial admixture risk to detecting pseudotaxa by mimicking categorical structures.

## Goals of the Present Review

The main goal of this systematic review is to critically examine the use of taxometric methods in the study of neurodevelopmental disorders (NDDs) within the published literature. We focused our review on the most commonly studied NDD categories: Autism Spectrum Disorder (ASD), Attention-Deficit/Hyperactivity Disorder (ADHD), and all forms of Specific Learning Disorders (SLDs), along with the closely related condition of Language Disorder/Impairment (the latter was not included in the preregistered protocol and thus constitutes a deviation). We excluded Intellectual Disability, as its etiological basis is often well identified and taxonic by definition in many cases (e.g., Down syndrome), or associated with identifiable medical risk factors. Residual non-syndromic cases, in turn, are thought to reflect the lower tail of general mental ability, and thus as inherently dimensional.

A core aim of the review is to critically assess the methodological quality and credibility of findings reported in published taxometric studies. We evaluated potential sources of bias such as artificial admixture, and whether sufficiently complete methodological information was reported to judge other threats to validity, including low indicator validity, nuisance covariance, and the skewness of indicators. (The latter is relevant when scales assess rare or extreme behaviors, and can create the illusion of categorical structure).

Following this systematic review, we present a Monte Carlo simulation study to empirically demonstrate the extent to which artificial admixture can lead to false detection of categorical latent structures in dimensional data, both when using taxometric analysis and cluster analysis.

# Systematic Review of Taxometric Analysis on NDDs

## Methods

The systematic review was conducted in accordance with the above listed goals, the preregistration protocol, and the PRISMA guidelines. The methods are detailed below.

### Sources and Search strategy

A systematic search was conducted on four databases: PsycINFO, PubMed, Scopus, Web of Science. No time limits were set, however virtually no records from before 1990 were expected, as taxometric methods were not yet developed or diffuse then. The following search query was performed on titles, abstracts, and/or keywords:

((“taxometric\*” OR “MAMBAC” OR “MAXCOV” OR “MAXEIG” OR “L-Mode”) AND (“autis\*” OR “ASD” OR “attention-deficit\*” OR “ADHD” OR “learning dis\*” OR “reading dis\*” OR “dyslexia” OR “math\* dis\*" OR "math\* diff\*"))

Where the wildcard "\*" was not supported (i.e., PubMed) it was replaced with full terms ("disorder", "disability" instead of "dis\*"; "mathematical", "mathematics", "math" instead of "math\*").

Language disorder was subsequently added as a deviation from preregistration, replacing the second part of the above search query with ("language imp\*" OR "language dis\*"). In addition, the full search query was launched on Google Scholar and the first 10 pages were screened; the references of and citations to @haslam2020metaanalysis were searched on Google Scholar for additional results; and a deep research via OpenAI's gpt-4o was launched providing it the preregistration protocol as context. All promising records thus emerging and not previously identified via databases, were listed under the *"Identification of studies via other methods"* section of the PRISMA flowchart.

### Eligibility Criteria

Inclusion criteria:

-   The study applies at least one taxometric procedure (e.g., MAMBAC, MAXCOV, L-Mode, MAXEIG);

-   The target population is individuals with NDDs or related traits;

-   The study draws a conclusion about the latent structure (categorical vs. dimensional) or reports sufficient information for an independent reader to reach a conclusion based on those results (e.g., reports CCFIs).

Exclusion criteria:

-   Theoretical, methodological, or simulation articles without empirical data;

-   Articles not reporting any behavioral data (e.g., only neuroimaging or neurophysiological data);

-   Reviews and/or meta-analyses;

-   Studies using other latent structure methods only (e.g., LPA, mixture modeling) without any taxometric methods being employed.

### Screening and Coding

Two independent human coders (MC and RP) screened and coded all articles. All discrepancies were resolved through discussion, and later revised by the PI (ET). No AI tools were used in the process, except in occasional cases by the human coders to assist them clarify the content of articles.

Since a single article could report multiple (sub)samples and/or multiple taxometric analyses on the same sample, the full coded dataframe followed a long-form data structure, with some articles repeating over multiple rows. Coded variables are listed in the data dictionary table reported in the preregistration protocol, to which a few variables for notes and coders agreement were added.

## Results

# Simulation of Artificial Admixture

We focus on artificial admixture because it is a very powerful threat to validity that could relatively easily be avoided by design when planning a study, and because as suggested by the result of the systematic review presented below, it is unfortunately frequent. There are different types of admixtures; @ruscio2004checklist discusses three cases. First, as mentioned above, the situation where clinical and nonclinical cases are separately recruited and merged. Second, the case of splitting a sample into subsamples (using alternative criteria) and then using those to test different taxometric questions. Third, trimming away observations of the putative complement (i.e., individuals without the disorder) to increase the base rate. All of these procedures artificially impose categorical structures to the data that do not exist in the population. It is worth mentioning that artificial admixture is also a threat for any type of cluster analysis, for the exact same reasons that it is a problem for taxometric analysis.

We simulate a dimensional scenario that makes it visually compelling how artificial admixture of the first type mentioned above creates pseudotaxa. A single normally-distributed latent variable $X$ is generated, that defines the target condition. A series of observed indicators $x1, x2, etc.$ are simulated, incorporating the $X$ latent variable plus some random normally-distributed noise ($R^2 = 0.77$). Individuals above the 99^th^ percentile of $X$ receive a diagnosis via clinical assessment. A sample of $N = 1000$ individuals is collected from the general population. Since the base rate is low (on average, only about $n = 10$ cases are expected to naturally present a diagnosis), researchers select and recruit an additional $n = 1000$ cases with a clinical diagnosis from the general population, thus creating artificial admixture. All data simulation is available in the online materials at xxxxxxxxxxxxxxxx.

An instance of the simulation, plotting two observed variables for illustrative purposes, is shown in @FIGURE_xx. Color is added for further clarity to distinguish individuals with and without a diagnosis, but two clusters still clearly emerge. Indicator validity is good (about $Cohen's$ $d = 2$), while nuisance covariance is partly problematic: for the clinical cases $r \approx 0$, but for the nonclinical cases $r = 0.61$. It is worth reminding that no taxa exist here, and the appearance of a categorical structure is mainly imposed by admixture, as shown below.

A Monte Carlo simulation was then performed, with 1000 iterations. Taxometric analysis was conducted using the "RTaxometric" package of R. As recommended, CCFI was used for inference: CCFI varies between 0 (maximal evidence for dimensional structure) to 1 (maximal evidence for categorical structure), with 0.5 indicating inconclusive results. Performed procedures were, as per default: MAMBAC, MAXEIG, and L-Mode. A third $x3$ observed indicator (similar to those above) was entered to allow conducting all these taxometric procedures. Over 1000 simulated datasets, results always strongly favored a categorical structure: average combined CCFI = xxxx, 95% quantile interval \[xxxx, xxxx\] (average MAMBAC CCFI = xxxx; average MAXEIG CCFI= xxxx; average L-Mode CCFI = xxxx). The smallest combined CCFI over 1000 iterations was xxxx. To confirm that artificial admixture also leads to misleading results when performing clustering, Gaussian Mixture Models (GMM) was also run testing 1 vs 2 components, using the "mclust" package of R: over 1000 iterations, results always favored the 2-component relative to the 1-component model, with an average $\Delta BIC = xxxx$ in favor of the 2-component model.

To confirm that the categorical structure was incorrectly favored specifically because of artificial admixture, we run another Monte Carlo simulation like the one above, but without admixture. In this case, a single sample of $N = 2000$ observations was drawn at each iteration. Therefore, nuisance covariance remains strong, but there is no admixture. Average combined CCFI = xxxx, 95% quantile interval \[xxxx, xxxx\]; average MAMBAC CCFI = xxxx; average MAXEIG CCFI= xxxx; average L-Mode CCFI = xxxx. Largest combined CCFI was xxxx. Thus, results remain frequently around inconclusive, probably due to nuisance covariance, but never strongly favor a categorical structure. GMM also consistently favored the 1-component model (over the 2-component model), with an average $\Delta BIC = xxxx$ in favor of the 1-component model. Therefore, admixture can be an extremely strong driver of false categorical findings, especially when many clinical cases are separately recruited and added to the sample.

@@

The present review suggests that taxometric analysis has been underused in neurodevelopmental research, despite its unique relevance for the recent debate on the categorical vs dimensional nature of disorders [@astle2022trandiagnostic; @michelini2024world; @happe2021both]. Focusing on selected NNDs, we identified 14 relevant records. The underuse of taxometric analysis in this field was expected: @haslam2020metaanalysis identified 183 articles employing taxometric methods on psychopathology, of which only 15 concerned childhood (these included various NNDs but also other conditions such as sleep problems). What is most worrying, many of the articles that we reviewed present serious methodological limitations and/or incomplete reporting of relevant information.

In the simulated example, artificial admixture was also accompanied by large nuisance covariance, although it was demonstrated that the latter posed a less severe problem to inference. In fact, the coexistence of low nuisance covariance and good indicator validity is practically impossible under a dimensional scenario. This is because an indicator can largely separate individuals with vs without a diagnosis if it is strongly related to the latent dimension, but this leads to nuisance covariance across valid indicators. Ultimately, however, such impasse might be interpreted as a direct signal that the true latent structure is dimensional.
